Manuscript: Detecting differences in Size Spectra

### 2023-02-23 Manuscript.Rmd
pre wesner pull commit

1.  The distribution of body size in communities  is remarkably consistent across habitats and taxa and can be represented by size spectra relationships.
2.  Many methods have been proposed for constructing size spectra relationships, but recent work has shown that varied methods return biased estimates of relationship parameters, and different methods in fact are not estimating the same parameter. Despite this variability in estimates, it is unclear if the relative change across environmental gradients is consistent across methodologies. Here, we simulate data sets across an hypothetical environmental gradient and estimate the size spectra parameter (slope or exponent, depending on method) at each site, and, importantly, estimate the relationship of parameters across the hypothetical gradient. We also use two previously published body size datasets across an anthropogenic stress gradient and an environmental temperature gradient and assess how the conclusions of these studies would vary based on methods used.
3.  We find that maximum likelihood methods always perform better than common binning methods. Additionally, the variance in estimates using MLE methods is markedly reduced when compared to binning methods.
4.  The uncertainty and variation in estimates when using binning methods is often greater than or equal to the variation previously published in experimental and observational studies, bringing into question the effect size of previously published results. However, re-analysis of two previously published datasets does not markedly alter the conclusions reached. Further study is needed to identify how and when these estimates may be biased, and when the general pattern of results can be considered to be "true".

@@

Body size distributions are a fundamental characteristic of communities. In general, size-abundance relationships are expected to scale with the value of -0.75 across all biological communities (Damuth 1981, 1991, 1998). This is thought to be a consequence of simple size-dependent metabolic constraints on organisms’ energy use predicted by the metabolic theory of ecology (Nee 1991, Brown et al. 2004). The remarkable consistency of these relationships across spatiotemporal scales and ecosystems has led them to be recommended as a "universal" indicator of ecological status [@Petchey2010]. Variation in size-abundance relationships have been documented through space [@pomeranz2022], time [@evans2022], seasonality [@mcgarvey2018], to human activities [@jennings2004]. Likewise, variation in size spectra relationships have been used to explain fundamental differences in how communities are organized. For example, external resource subsidies "bend the rules" and allow higher abundances of large body sizes than would be expected [@perkins2018]. However, recent research has shown that these results may be an artifact of how the data were treated. Edwards et al. (2020) reanalyzed a time series of marine fisheries data and found that previously reported changes through time were actually dependent on the methodology used.

Individual size distributions (ISD *sensu* [@White2007], also referred to as abundance size spectra) are one of the size-abundance relationships commonly used. ISDs show the relationship between the abundances of individual body size regardless of taxonomy. Generally, there is a negative relationship between individual body size (M) on the *x*-axis  and abundance (N) on the *y*-axis. Theoretical and empirical data support this relationship being described as a simple power law with exponent $\lambda$ in the form of $N \sim M^{\lambda}$, where $\lambda = 2$ [@Sheldon1972, @Andersen2006a]. Commonly, $N$ is the count of body sizes grouped into bins, and $\lambda$ is estimated as the slope from OLS regressions of log transformed data of $N$ and the mid point of the body size bin $M_{bin}$: $\log(N_{count}) = \lambda \log(M_{bin})$. Myriad binning methods have been proposed, including linear and logarithmic bin widths. Likewise, methods can rely on the absolute count in the bins as well as normalization techniques where the count is divided by the bin width (especially common with logarithmic binning). Alternatively, $\lambda$ can be estimated directly using maximum likelihood techniques.

Previous work has shown that the estimates of $\lambda$ differ between MLE and OLS techniques. OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2008, Edwards et al. 2017), and reanalysis of empirical data also indicates that the conclusions are dependent on the methodology used [@white2008 @edwards2020]. However, recent empirical analysis of stream macroinvertebrate communities across the National Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied based on method, the relative change across the environmental gradient was consistent across methods [@pomeranz2022]. While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of OLS slope parameters of 0.1 coincide with a relative change of MLE $\lambda$ estimates of 0.1?

We had three primary objectives in this study: 1) how well do different methods estimate site-specific $\lambda$'s? 2) Do the estimates of the hypothetical relationship obtained from different methods agree? and 3) Are the conclusions reached from previously published data sets dependent on the method used?

In order to answer the first two questions, we simulate body size observations from bounded power law distributions with varied $\lambda$ exponents across a hypothetical environmental gradient and compare the results obtained using three different methodologies common in the literature. To answer the third question, we re-analyze two previously published data sets of community body sizes across a stress and environmental gradient, respectively [@pomeranz2019, @pomeranz2022]. We find that the MLE method more accurately estimates the site-specific $\lambda$ exponents as well as the relative change in exponents across the hypothetical gradient, as well as having smaller variation around the estimates. The logarithmic binning methods generally perform well, but have larger variation around the site-specific estimates. In addition, the variation of the estimates of the hypothetical relationship obtained from the binning methods is dependent on the value of $\lambda$ used (under and over estimates steep and shallow $\lambda$, respectively).

@@

In order to investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in [@edwards2017]. Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values are based on empirical body sizes of stream benthic communities reported in [@pomeranz2022]. Our results are not dependent on the range of body sizes, and we show the results of other body size ranges in the supplemental information.

@@

We sampled $n = 1000$ body sizes from distributions described by five different $\lambda$'s: (-1.5, -1.75, -2.00, -2.25, -2.5). The values of $\lambda$ describe how quickly the abundance of large body sizes decline within a community, e.g., a value of -1.5 means there would be a relatively high number of large body sizes (shallow) whereas a value of -2.5 means there would be relatively very few large body sizes (steep). We used three different methods (described below) to estimate the value of $\lambda$. We repeated this process 1000 times (reps) and plot the distribution of values obtained for each method.

@@

For each sample of $M$ and each of the 1000 replicates, we estimated the exponent $\lambda$ directly using MLE methods modified from the the `sizeSpectra` package (Edwards et al. 2020). Throughout the manuscript, these estimates are referred to as MLE.

@@

In addition, we use two common binning methods to estimate the OLS slope parameter in log-log space. It is important to note that the binning methods are estimating a slope ($\beta_{1}$) from an OLS regression analysis, and not $\lambda$ directly. However, in the interest of clarity, we will refer to all estimates as $\lambda$ regardless of the method used, and reserve the term $\beta_{env}$ to refer to the estimate of the size spectra relationships across the hypothetical environmental gradient.

@@

### Log~2~ Bins: NAS
The second method was similar to ELBn, but bins of Log~2~ widths are used. The number of bins is dependent on the range of body sizes present in the data. When working with empirical data with different size ranges this can alter the number of bins per site, and it is recommended to construct the bins based on the global min and max body size present. However, since the data here were simulated from a known size range, the number of bins for each site is identical. The count in each bin is normalized in the same way as described above. This method has been  used [@jennings2004, @Jennings2002, @sprules2015, @pomeranz2019; @mcgarvey2018) and is referred to as the Normalized Abundance Spectrum (NAS).

@@

Where $\beta_{1}$ is considered to be the $\lambda_{estimate}$.

@@

### Hypothetical relationshiop of -0.5

@@

To do this, we again sampled body sizes from distributions described by five different $\lambda$'s. However, because the estimates from the binning methods varied depending on the value of $\lambda$ (see results), we chose three different "windows" of lambda values. The three windows were steep ($\lambda = [-1.50,~ -1.75,~ -2.00,~ -2.25, -2.50]$), medium ($\lambda = [-1.00,~ -1.25,~ -1.50,~ -1.75,~ -2.00]$), and shallow (($\lambda = [-0.50,~ -0.75,~ -1.00,~ -1.25,~ -1.50]$).

**NOTE** This MS seems to be getting long. Possibly present "medium" window in main results, and include other windows in SI

Each sample of body sizes from different $\lambda$ values was assumed to come from a community across a hypothetical environmental gradient $X$. Values of $X$ for each community are uniformly distributed from -1 to 1 (site-specific x-values: $(-1.0,~-0.5,~0,~0.5,~1.0)$. For each $\lambda$ "window", the relative change in $\lambda$ values ($\beta_{env}$) across the gradient is equal to -0.5, such that our model is:

**JEFF** Is this correct? for some reason this seems wrong to me.

$$\beta_{env} = Norm(\alpha -0.5X,~\sigma)$$

We repeated the data simulation 1000 times (reps). The main results presented here were not dependent on the range of *x*-values or the number of sites (supplemental information).

### Varying Hypothetical Relationship

Finally, we also varied the relationship ($\beta_{env}$) across the hypothetical gradient. We tested $\beta_{env}$ values of 0, -0.25, and -0.5 by sampling 5 body sizes communities uniformly distributed across the following sets of $\lambda$ values: [-1.5], [-1.25, -1.75], [-1, -2].

## Estimation of relationship across gradient, $\beta_{env}$

For each simulation replicate and size spectra method, we estimated how the size spectra parameters varied across the hypothetical gradient. We did this by performing simple OLS regression of the $\lambda_{estimate}$ across the hypothetical gradient ($X$):

$$\lambda_{estimate} = \beta_0 + \beta_{env} * X$$

The distribution of the relationship coefficient, $\beta_{env}$, were plotted compared to the known relationship.

@@

We re-analyze two data sets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage (AMD) stress gradient. Details of the sample collection and processing can be found in [@pomeranz2019]. Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using taxon-specific published length weight regressions.

@@

>>>>

1.  The distribution of body size in communities  is remarkably consistent across habitats and taxa and can be represented by size spectra, which are described by a power law. The focus of size spectra analysis is to estimate the exponent ($\lambda$) of the power law.
2.  Many methods have been proposed for estimating $\lambda$ most of which involve binning the data, summing abundance within bins, and then fitting a ordinary least squares (OLS) regression in log-log space. However, recent work has shown that binning procedures may return biased estimates of size spectra exponents compared to procedures that directly estimate $\lambda$ using maximum likelihood estimation (MLE). Despite this variability in estimates, it is unclear if the relative change across environmental gradients is consistent across methodologies. Here, we used simulation to compare the ability of two binning methods (NAS, ELBn) and MLE to 1) recapture known values of $\lambda$, and 2) recapture parameters in a linear regression measuring the change in $\lambda$ across a hypothetical environmental gradient. We also compared the methods using two previously published body size datasets across a pollution gradient and a temperature gradient 
3.  Maximum likelihood methods always performed better than common binning methods, which demonstrated consistent bias depending on the simulated values of $\lambda$. This bias carried over to the regressions, which were more accurate when $\lambda$ was estimated using MLE compared to the binning procedures. Additionally, the variance in estimates using MLE methods is markedly reduced when compared to binning methods.
4.  The uncertainty and variation in estimates when using binning methods is often greater than or equal to the variation previously published in experimental and observational studies, bringing into question the effect size of previously published results. However, while the methods produced different slope estimates from previously published datasets, they were in qualitative agreement on the sign of those slopes (i.e., all negative or all positive). Our results provide further support for the direct estimation of $\lambda$ using MLE (or similar procedures) over the more common methods of binning.

@@

Body size distributions are a fundamental characteristic of communities. In general, size-abundance relationships are expected to scale with the value of -0.75 across all biological communities [@Damuth1981, @Damuth1991, @Damuth1998]. This is thought to be a consequence of simple size-dependent metabolic constraints on organisms’ energy use predicted by the metabolic theory of ecology [@Nee1991, @Brown2004]. The remarkable consistency of these relationships across spatiotemporal scales and ecosystems has led them to be recommended as a "universal" indicator of ecological status [@Petchey2010]. Variation in size-abundance relationships have been documented through space [@pomeranz2022], time [@evans2022], seasonality [@mcgarvey2018], and in response to human activities [@jennings2004]. Likewise, variation in size spectra relationships have been used to explain fundamental differences in how communities are organized. For example, external resource subsidies "bend the rules" and allow higher abundances of large body sizes than would be expected [@perkins2018]. However, recent research has shown that these results may be an artifact of how the data were treated. Edwards et al. (2020) reanalyzed a time series of marine fisheries data and found that previously reported changes through time were actually dependent on the methodology used.

Individual size distributions (ISD *sensu* [@White2007], also referred to as abundance size spectra) are one of the size-abundance relationships commonly used. ISDs represent frequency distributions of individual body sizes within a community, regardless of taxonomy. Generally, there is a negative relationship between individual body size (M) on the *x*-axis  and abundance (N) on the *y*-axis. Theoretical and empirical data support this relationship being described as a simple power law with exponent $\lambda$ in the form of $N \sim M^{\lambda}$, where $\lambda = 2$ [@Sheldon1972, @Andersen2006a]. Commonly, $N$ is the count of body sizes grouped into bins, and $\lambda$ is estimated as the slope from OLS regressions in log-log space of $N$ and the mid point of the body size bin $M_{bin}$: $\log(N_{count}) = \lambda \log(M_{bin})$. Myriad binning methods have been proposed, including linear and logarithmic bin widths. Likewise, some methods rely on the absolute counts in the bins and others employ normalization techniques such as dividing the count by the bin width (especially common with logarithmic binning). As an alternative to binning methods, $\lambda$ can be estimated directly on un-binned data using maximum likelihood techniques.

Previous work has shown that the estimates of $\lambda$ differ between MLE and size-binned OLS techniques. Size-binned OLS methods are particularly sensitive to decisions made in the binning process. Simulation studies have shown that MLE offers consistently more accurate estimates of $\lambda$ (White et al. 2008, Edwards et al. 2017), and reanalysis of empirical data also indicates that the conclusions are dependent on the methodology used [@white2008 @edwards2020]. However, recent empirical analysis of stream macroinvertebrate communities across the National Ecological Observatory Network (NEON, USA) showed that while the estimates of $\lambda$ varied based on method, the relative change across the environmental gradient was consistent across methods [@pomeranz2022]. While there is a growing consensus that MLE methods offer more reliable estimates of $\lambda$, and binning methods result in biased estimates, it remains unclear if these biases are consistent and systematic or stochastic, and whether or not the relative change in ISD parameters is consistent across space and time. In other words, if the data within a study are all treated the same, does a relative change of size-binned OLS slope parameters of 0.1 coincide with a relative change of MLE $\lambda$ estimates of 0.1?

We had Two primary objectives in this study: 1) to compare how well different methods estimate site-specific $\lambda$'s 2) To compare how different methods of estimating $\lambda$ impact subsequent regressions that use $\lambda$'s as a response variable. We did this using both simulated and empirical data sets.

@@

To investigate the performance of commonly used methods, we simulate body size observations from a bounded power law distribution using the inverse method, as described in [@edwards2017]. Let $M$ be a random variable of body sizes described by the probability density function:

@@

$M_{min} = 0.0026$ and $M_{max} = 1.2 *10^3$. These values were based on empirical body sizes of stream benthic communities reported in [@pomeranz2022]. Our results were not dependent on the range of body sizes (Supplemental Information).

@@

Using the procedure above, we simulated $n = 1000$ body sizes from five different $\lambda$'s: (-1.5, -1.75, -2.00, -2.25, -2.5). The values of $\lambda$ describe how quickly the abundance of large body sizes decline within a community. For example, a value of -1.5 means there would be a relatively high number of large body sizes (shallow) whereas a value of -2.5 means there would be relatively very few large body sizes (steep).

@@

After simulating data, we used three different methods (described below) to estimate the value of $\lambda$ (maximum likelihood, equal logarithmic bins, and normalized abundance spectra). We repeated this process 1000 times (reps) and plotted the distribution of values obtained for each method.

## Maximum likelihood estimation (MLE)

We estimated the exponent $\lambda$ directly using MLE methods modified from the the `sizeSpectra` package [@edwards2017]. Throughout the manuscript, these estimates are referred to as MLE.

@@

In addition, to estimate the OLS slope parameter in log-log space, we used two common binning methods. . It is important to note that the log-log regression is estimating $\lambda$ as the regressionslope ($\beta_{1}$), and not $\lambda$ directly. In the interest of clarity, we will refer to all estimates of lambda as $\lambda$ regardless of the method used, and reserve the terms “parameters”, “intercept”, and “slope” $\beta_{env}$ to refer to the parameters of linear regression that uses the lambdas as a response variable.

@@

### Log~2~ Bins: NAS
The second method was similar to ELBn, but bins of log~2~ widths are used. The number of bins is dependent on the range of body sizes present in the data. When working with empirical data with different size ranges this can alter the number of bins per site, and it is recommended to construct the bins based on the global min and max body size present. However, since the data here were simulated from a known size range, the number of bins for each site is identical. The count in each bin is normalized in the same way as described above. This method has been  used extensively in the literature [\e.g., @jennings2004, @Jennings2002, @sprules2015, @pomeranz2019; @mcgarvey2018) and is referred to as the Normalized Abundance Spectrum (NAS).

@@

Where $N$ is the count in each bin, $M$ is the mid-point of each bin, and $\beta_{1}$ is considered to be the $\lambda_{estimate}$.

@@

@@

To do this, we first simulated  $\lambda$'s from a linear regression with a known slope $\beta_{1}$ of -0.5 and intercept of -1.5 across a generic predictor variable x with values ranging from -1 to 1:

$$\lambda_{ik} = \beta_{0} + \beta_{0}x_{ik}$$

This produced $k$ = 1000  values from each value of $x$ (-1, -0.5, 0, 0.5, 1), resulting in 5000 total $\lambda$’s. From each $\lambda$ we simulated 1000 individual body sizes using the procedure described above.

Once the body size data were simulated, we then reversed the procedure. First, we used the three methods (NAS, ELBn, MLE) to estimate $\lambda$ for each of the 5000 body size datasets. Then we fit a linear regression to those data with the predictor variable $x$ and compared the resulting intercept ($\beta_{0}$) and slope ($\beta_{1}$) parameter estimates to the known value. The main results presented here were not dependent on the range of x-values or the number of sites (Supplemental Information).

@@

We re-analyzed two data sets of benthic macroinvertebrate communities from stream habitats across two different gradients. In the first, quantitative macroinvertebrate samples were collected from streams across an acid mine drainage (AMD) stress gradient. Details of the sample collection and processing can be found in [@pomeranz2019]. Briefly, all individuals from each sample were identified to the lowest practical taxonomic unit and body lengths were measured using image processing software from photos taken with a camera mounted to a dissecting microscope. Body mass was estimated using taxon-specific published length weight regressions.

@@

## Performance Metrics

We compared performance of each procedure (NAS, ELBn, MLE) by first plotting the distribution of lambda or regression parameter estimates ($\lambda$, $\beta_{0}$, $\beta_{1}$) against the known values. We calculated bias for the procedures overall as the median absolute difference (averaged across all simulations) between the known values and the modeled estimates. Finally, for each procedure we estimated the width of the 95% CI’s to compare uncertainty.
